---
title: "PCIS+ Analysis: Distributions and counts"
format:
  html:
    df-print: paged
    self-contained: true
    prefer-html: true
---

```{r}
#| echo: false
#| output: false
library(tidyverse)
library(arrow)

fmt_comma <- scales::label_comma(accuracy = 1)
fmt_percent <- scales::label_percent()

count_prop <- function(df, ...) {
  df %>%
    count(...) %>%
    mutate(prop = round(n / sum(n), 2))
}

fmt_count_prop <- function(df) {
  df %>%
    mutate(
      n = fmt_comma(n),
      prop = fmt_percent(prop)
    )
}

positions <- read_parquet("../data/out/positions.parquet")
positions_binned <- read_parquet("../data/out/positions-binned.parquet")

positions_n_organizations <- positions %>% distinct(organization) %>% pull %>% length
```

There are `r positions %>% count %>% pull %>% fmt_comma` positions (better known as “boxes”) across `r positions_n_organizations` organizations (which I’ll sometimes call “departments”, even if that’s not quite accurate) in the PCIS+ database. Here’s how they break down between “occupied”—i.e., a real person has the position (sits in the box, to carry on with the metaphor)—and vacant:

```{r}
#| echo: false
positions %>%
  count_prop(position_status) %>%
  fmt_count_prop
```

How does this break across departments? Which departments have the most?

```{r}
#| echo: false
positions %>%
  count_prop(organization) %>%
  arrange(-n) %>%
  fmt_count_prop
```

And the least?

```{r}
#| echo: false
positions %>%
  count_prop(organization) %>%
  arrange(n) %>%
  fmt_count_prop
```

Depending on how closely you follow federal public administrivia, the counts for the largest departments may surprise you, versus [their actual sizes in terms of headcount](https://www.canada.ca/en/treasury-board-secretariat/services/innovation/human-resources-statistics/population-federal-public-service-department.html)—these reveal some important characteristics of this dataset to keep in mind when using it for analysis:

- PCIS+ tracks the number of _boxes_ a department has created—oftentimes, a department will have many more (up to 50% more, from a quick scan of the data) positions than they will people; this isn’t necessarily an indication of how large a department plans to grow, so much as a reflection of incentives in HR policy and processes (to vastly oversimplify, it’s a lot of work to classify a position, so it’s easier to keep it around than to cut out unused boxes)
- PCIS+ only includes positions in [the core public administration—i.e., organizations with the Treasury Board as the employer](https://www.canada.ca/en/public-service-commission/services/public-service-hiring-guides/information-staffing-obligations.html)—significant organizations like the Canada Revenue Agency, as well as, e.g., service personnel in the Armed Forces and the RCMP, are not included

If we instead focus on _occupied_ positions (i.e., the number of people, roughly), the figures get closer to what we find elsewhere:

```{r}
#| echo: false
positions %>%
  filter(position_status == "Occupied") %>%
  count_prop(organization) %>%
  arrange(-n) %>%
  fmt_count_prop
```

Discrepancies can be simplified as “counting people is hard”. To go one level deeper, though:

- how quickly position data are updated in PCIS+ versus reality is unclear, and some people may be double counted (or a position classified as “occupied” even while its occupant is on long-term leave, which wouldn’t count in other data on employee numbers)
- [public data on employee numbers based on pay system data](https://www.canada.ca/en/treasury-board-secretariat/services/innovation/human-resources-statistics/population-federal-public-service-department.html) tends to be as of March 31, while the PCIS+ export powering this analysis is from around November 2022—there’s definitely seasonal variation in public service employment, depending on the organization, as well as general change over time
- these are headcounts, not full-time equivalent (FTE) counts—some (sometimes many, depending on the department) positions are part-time positions, often for seasonal posts, so while these count as 1 position, they may only be 0.3 FTE

So, with those caveats aside—let’s explore the data!



## Classification groups, generic and specific

```{r}
#| echo: false
#| output: false
classification_group_counts <- positions %>%
  group_by(group, group_label) %>%
  summarize(
    n_positions = n(),
    n_organizations = n_distinct(organization)
  ) %>%
  ungroup
```

Across the positions, there are some `r classification_group_counts %>% nrow %>% fmt_comma` different classification groups.

Some groups are very common, appearing in most departments—we’ll call them “generic”:

```{r}
#| echo: false
classification_group_counts %>%
  slice_max(n_organizations, n = 10) %>%
  arrange(-n_organizations, -n_positions)
```

From this, we see that AS, EX, and IT are the most common groups, found in almost every organization.

Some, on the other hand, are very rare, appearing in fewer than 25% of departments (`r floor(0.25 * positions_n_organizations)`)—we’ll call them “specific”:

```{r}
#| echo: false
classification_group_counts %>%
  filter(n_organizations < (0.25 * positions_n_organizations)) %>%
  slice_min(n_organizations, n = 10) %>%
  arrange(n_organizations, -n_positions)
```

These include a number of groups that appear only in one department—Border Services and Correctional Services groups, for example, are only found in CBSA and CSC respectively.

...which organizations have the highest number of specific groups? (probably DND—firefighting, defense science, teaching, etc) what's the largest organization to have only generic groups? (maybe we should redefine generic groups)
