---
title: "PCIS+ Analysis: Vacancies"
format:
  html:
    df-print: paged
    self-contained: true
    prefer-html: true
---

```{r}
#| echo: false
#| output: false
library(tidyverse)
library(arrow)

count_prop <- function(df, ...) {
  df %>%
    count(...) %>%
    mutate(prop = round(n / sum(n), 2))
}

positions <- read_parquet("../data/out/positions.parquet")
```

How many positions are occupied versus vacant?

```{r}
#| echo: false
positions %>%
  count_prop(position_status)
```

Which organizations have the highest vacancy rates? (The ones that really need to hire.)

```{r}
#| echo: false
positions %>%
  group_by(organization) %>%
  count_prop(position_status) %>%
  ungroup %>%
  filter(position_status == "Vacant") %>%
  select(-position_status) %>%
  slice_max(prop, n = 5)
```

Which have the lowest? (The ones that are really popular.)

```{r}
#| echo: false
positions %>%
  group_by(organization) %>%
  count_prop(position_status) %>%
  ungroup %>%
  filter(position_status == "Vacant") %>%
  select(-position_status) %>%
  slice_min(prop, n = 5)
```

Removing small organizations, which have the highest vacancy rates?

```{r}
#| echo: false
positions %>%
  filter(org_n_positions > 250) %>%
  group_by(organization) %>%
  count_prop(position_status) %>%
  ungroup %>%
  filter(position_status == "Vacant") %>%
  select(-position_status) %>%
  slice_max(prop, n = 5)
```

Removing small organizations, which have the lowest?

```{r}
#| echo: false
positions %>%
  filter(org_n_positions > 250) %>%
  group_by(organization) %>%
  count_prop(position_status) %>%
  ungroup %>%
  filter(position_status == "Vacant") %>%
  select(-position_status) %>%
  slice_min(prop, n = 5)
```


...Which departments (directorates?), classifications, levels have more vacant positions?
